Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-01 LICS21/antitone.tex

%\paragraph{}
As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.

Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

As a corollary, any term which admits an antitone partial ranking function terminates almost surely.

>>>>

%\paragraph{}
As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone sparse ranking function.

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

As a corollary, any term which admits an antitone sparse ranking function terminates almost surely.

### 2021-01-01 LICS21/ranking.tex

\subsection{Partial ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.

@@

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A partial ranking function that is total is just a ranking function. Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

\begin{theorem} \label{thm:partial implies rankable}
Every partial ranking function is a restriction of a ranking function.
\end{theorem}
\begin{proof}
Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions. Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
\lo{TODO. The preceding sentence deserves a proof.}
\end{proof}

As a corollary, any term which admits a partial ranking function terminates almost surely.

\subsection{Examples}
Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
\begin{align*}
E[M \oplus_p N] \to^3 & E[M] & \\
E[M \oplus_p N] \to^3 & E[N] & \\
\red^3(E[M \oplus_p N], s) = & \left\{
    \begin{array}{ll}
        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
    \end{array} \right .
\end{align*}
A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$. Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function. It is therefore possible to prove rankability directly using the shortcutted reductions.

A similar procedure would work for other forms of syntactic sugar. If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination. There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.

%\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a partial ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.

>>>>

\subsection{Sparse ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.

@@

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

\begin{theorem} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{theorem}
\begin{proof}
Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions. Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
\lo{TODO. The preceding sentence deserves a proof.}
\end{proof}

As a corollary, any term which admits a sparse ranking function terminates almost surely.

\subsection{Examples}
Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
\begin{align*}
E[M \oplus_p N] \to^3 & E[M] & \\
E[M \oplus_p N] \to^3 & E[N] & \\
\red^3(E[M \oplus_p N], s) = & \left\{
    \begin{array}{ll}
        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
    \end{array} \right .
\end{align*}
A sparse ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a sparse ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$. Such a pseudo-sparse ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual sparse ranking function. It is therefore possible to prove rankability directly using the shortcutted reductions.

A similar procedure would work for other forms of syntactic sugar. If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination. There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.

%\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a sparse ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a sparse ranking function for this term.
